---
title: "Lab 13"
author: "Victoria Means"
format: pdf
editor: visual
---

## Section 1. Differential Expression Analysis 

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

```
```{r}
# Import metadata and take a peak
colData = read.csv(metaFile, row.names=1)
head(colData)

# Import countdata
countData = read.csv(countFile, row.names=1)
head(countData)
```
Q1. Complete the code below to remove the troublesome first column from countData 
```{r}
# Note we need to remove the odd first $length col
countData <- as.matrix(countData[,-1])
head(countData)
```
Q2. Complete the code below to filter countData to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns)
```{r}
# Filter count data where you have 0 read count across all samples.
countData = countData[-rowSums(countData==0), ]
head(countData)

```
```{r}
library(DESeq2)
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
dds
```


```{r}
res = results(dds)
```

Q3. Call the summary() function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff.
```{r}
summary(res)
```
```{r}
plot( res$log2FoldChange, -log(res$padj) )
```
Q4. Improve the plot by completing the below code, which adds color and axis labels 
```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```
Q5. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below.
```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")
res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

head(res, 10)
```
Q6. Finally for this section let's reorder these results by adjusted p-value and save them to a CSV file in your current project directory.
```{r}
res = res[order(res$pvalue),]
write.csv(res, file="deseq_results.csv")
```
##Section 2. Pathway Analysis 

```{r}
library(pathview)
library(gage)
library(gageData)
```

```{r}
data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)

```
```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```
```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```
```{r}
attributes(keggres)
```
```{r}
# Look at the first few down (less) pathways
head(keggres$less)
```
```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```
```{r}
# A different PDF based output of the same data
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```
```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```
Q7.Can you do the same procedure as above to plot the pathview figures for the top 5 down-reguled pathways?
```{r}
attributes(keggres)
```
```{r}
head(keggres$greater)
head(keggres$less)
head(keggres$stats)
```
```{r}
lapply(keggres, head)
```

```{r}
pathways = data.frame(id=rownames(keggres$less), keggres$less)
head(pathways)
```
```{r}
keggrespathways.1 <- rownames(keggres$less)[1:5]

# Extract the 8 character long IDs part of each string
keggresids.1 = substr(keggrespathways.1, start=1, stop=8)
keggresids.1

pathview(gene.data=foldchanges, pathway.id=  keggresids.1, species="hsa")
```


## Section 3. Gene Ontology (GO)
```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

##Section 4. Reactome Analysis 
```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```
```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```
Q8. What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

Based on the "Entities p-value", the most significant pathway is the "Signaling by Rho GPTases, Miro GTPases and RHOBTB3". From the previous KEGG results, the most significant pathway was the cell-cycle. These differences are due to how the data is categorized, as one analyzes the pathway through the mapping of genes while the other is through the pathway of biological molecules and their relation to pathways and processes. 


